Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-10-26 manuscript.tex
edits

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y is more than 40-fold lower than on the X, suggesting an extensive reduction in diversity. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model that accounts for the occurrence of female-biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and population genetic estimates of the distribution of fitness effects of deleterious mutations, we show that Y diversity levels can be explained by the action of background selection. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study is consistent with the prediction that the early stages of Y chromosome degeneration are characterized by strong effects of selective interference experienced by a large number of selected sites. Although we do not exclude the possibility that positive selection has also reduced Y chromosome variability, our simulations suggest that the reduction in diversity arising from purifying selection is sufficient to explain our observed patters of neutral variation.

\end{abstract}

@@

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes. On the other hand, if gene silencing and Y degeneration occur very early during the process of Y chromosome degeneration, it is possible that the strength of linked selection is minimal even on relatively young Y chromosomes, and that much degeneration may occur neutrally following the silencing of Y-linked genes.

@@

With with no variation in mating success in either sex ($C^2_{m}=C^2_{f}=0$), equations 1-3 reduce to the familiar expressions from \citep{wright1931evolution}:

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y is more than 40-fold lower than on the X, suggesting an extensive reduction in diversity. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model that accounts for the occurrence of female-biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and population genetic estimates of the distribution of fitness effects of deleterious mutations, we show that Y diversity levels can be explained by the action of linked selection against deleterious mutations, although selective sweeps may also be playing a role. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study is consistent with the prediction that the early stages of Y chromosome degeneration are characterized by strong effects of selective interference experienced by a large number of selected sites. These results suggest that selective interference is playing a major role in the ongoing degeneration of the Y chromosome, rather than simply neutral drift following gene silencing. 
\end{abstract}

@@

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes. On the other hand, if gene silencing and Y inactivation occur very early during the process of Y chromosome degeneration, it is possible that the strength of linked selection is minimal even on relatively young Y chromosomes, and that much degeneration may occur neutrally following the silencing of Y-linked genes.

@@

With no variation in mating success in either sex ($C^2_{m}=C^2_{f}=0$), equations 1-3 reduce to the familiar expressions from \citep{wright1931evolution}:
